---
title: 'Assignment 3: Data Exploration'
author: "Erin Lee"
date: "Spring 2026"
output:
  word_document: default
  pdf_document: default
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  [NEW] Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to Canvas.

8.  Initial here to acknowledge that you did not use AI in completing this assignment, except where expressly allowed: E.L

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks in your code chunks.

**TIP**: If your code fails to knit, check:
 * That no `install.packages()` or `View()` commands exist in your code. 
 * That you are not displaying the entire contents of a large dataframe in your code.

---

## Set up your R session

1.  Load necessary packages (tidyverse, here), check your current working directory and import two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 

**Be sure to**:
* Use the `here()` package in specifying the paths to your datasets
* Include the appropriate subcommand to read in character based columns as factors


```{r, setup, message = FALSE}

# load packages
library(tidyverse)
library(here)

# check workding directory
here()

# read data and convert character columns to factors
Neonics <- read.csv(
  file = here('Data', 'Raw', 'ECOTOX_Neonicotinoids_Insects_raw.csv'),
  stringsAsFactors = TRUE
)

Litter <- read.csv(
  file = here('Data', 'Raw', 'NEON_NIWO_Litter_massdata_2018-08_raw.csv'),
  stringsAsFactors = TRUE
)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. (AI is allowed here, but put answers in your own words.)

> Answer: Insects play a critical role in maintaining ecosystems through services such as pollination and natural pest control. Harmful effects of neonicotinoids on insect nervous systems can reduce these functions, leading to decreased crop productivity, loss of biodiversity, and cascading impacts on other organisms through the food web.


3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. (AI is allowed here, but put answers in your own words.)

> Answer: Forest litter and woody debris are major sources of organic matter that supply nutrients and carbon to the soil, influencing microbial activity, soil fertility, moisture retention, and carbon storage. Studying their decomposition rates and accumulation helps assess forest ecosystem health and long-term carbon cycling, as well as ecosystem responses to climate and environmental change.


4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter and fine woody debris are collected using different types of traps
 2. Samples are classified as litter or fine woody debris based on size criteria
 3. Sampling frequency varies depending on trap type and vegetation type


## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?
> Answer: The dataset has 4,623 rows and 30 columns.

```{r, dimensions}
# find the dimensions of the dataset
dim(Neonics) 

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied.  [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r, summary1}
# determine the most common effects studied
sort(summary(Neonics$Effect), decreasing = TRUE)

```

► Question: Which two effects stand out as the most studied? Can you guess why these effects might specifically be of interest?
> Answer: Population and Morality. These effects are of particular interest because they directly reflect impacts on survival and population-level dynamics, which are critical for assessing ecological risk and long-term consequences of neonicotinoid exposure on insect populations.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name).[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r, summary2}
# determine the six most commonly studied species
summary(Neonics$Species.Common.Name, maxsum = 7)

```

► Question: What do these species have in common? Why might they be of interest over other insects? 
> Answer: The six most commonly studied species are honey bee (667), parasitic wasp (285), buff tailed bumblebee (183), Carniolan honey bee (152), bumble bee (140), and Italian honey bee (113), all of which are bees or closely related hymenopterans. These species are of particular interest because they play key roles as pollinators or beneficial insects in ecosystems and agriculture, and neonicotinoids are known to strongly affect their nervous systems, making them important indicators of ecological and agricultural risk.


8.  The `Conc.1..Author` column, which lists the concentration of the neonicitoid dose, should include numeric values. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r, class}
# check the class of Con.1..Author column
class(Neonics$Conc.1..Author.)
#View(Neonics)

```

> Answer: This is because the column contains non-numeric values such as units, ranges, or symbols (e.g., “<” or “>”), which prevent R from interpreting the data as purely numeric.


## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r, visulaization-freqpoly(1), fig.height = 3, fig.width = 4}

# check the minimum and maximum value of the data
min(Neonics$Publication.Year)
max(Neonics$Publication.Year)
unique(Neonics$Publication.Year)
# generate a frequency line graph using 5-year bins with publication years limited to 1980–2020
ggplot(Neonics) + 
geom_freqpoly(aes(x = Publication.Year), binwidth = 5) +
  scale_x_continuous(limits = c(1980, 2020))


```


10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r, visulaization-freqpoly(2), fig.height = 3, fig.width = 4}
# generate a frequency line graph using 5-year bins (1980–2020), with different Test.Location values shown in different colors
ggplot(Neonics) + 
geom_freqpoly(aes(x = Publication.Year, color=Test.Location), binwidth = 5) +
  scale_x_continuous(limits = c(1980, 2020))

```


►Interpret this graph. What are the most common test locations, and do they differ over time?
> Answer: aboratory studies are the most common test location and become increasingly dominant over time, especially after the early 2000s. Field natural studies are the second most common and show temporal variation, while field artificial and undeterminable locations remain rare throughout the period.


11. Create a bar graph of Endpoint counts. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r, visulaization-bar graph, fig.height = 3, fig.width = 4}
# generate a bar graph
ggplot(data = Neonics, aes(x = Endpoint)) + 
geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```


► What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix (p.721) for more information. 
> Answer: The two most common endpoints are NOEL and LOEL. NOEL (No-Observable-Effect Level) is defined as the highest dose or concentration at which no statistically significant effects are observed compared to the control, while LOEL (Lowest-Observable-Effect Level) is the lowest dose or concentration that produces effects significantly different from the control. Both endpoints are widely used in terrestrial ecotoxicology to assess threshold levels of chemical toxicity.

---

## Explore your data (Litter)

12. Determine the class of `collectDate`. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.
> Answer: "2018-08-02" "2018-08-30"

```{r, date}
# check the class of 'collectDate'
class(Litter$collectDate)

# change to a date 
Litter$collectDate <- as.Date(Litter$collectDate)
class(Litter$collectDate)

# determine the dates litter was sampled in August 2018
unique(
  Litter$collectDate[
    Litter$collectDate >= as.Date("2018-08-01") &
    Litter$collectDate <= as.Date("2018-08-31")
  ]
)

```


13. Using the `unique` function, list the different `plotIDs` sampled at Niwot Ridge. 
> Answer: 12 plot IDs - NIWO_061 NIWO_064 NIWO_067 NIWO_040 NIWO_041 NIWO_063 NIWO_047 NIWO_051 NIWO_058 NIWO_046 NIWO_062 NIWO_057

```{r, list of plotIDs}
unique(Litter$plotID)

```

► How is the information obtained from `unique` different from that obtained from `summary`?
> Answer: The unique function returns a list of all distinct values present in a variable, showing exactly which plotIDs were sampled. In contrast, the summary function provides frequency information for each value, indicating how many times each plotID appears in the dataset rather than simply listing the unique identifiers.


14. Create a bar graph of `functionalGroup` counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
> Answer: The bar graph shows that the different litter functional groups are relatively evenly distributed across the Niwot Ridge sites, with no single litter type overwhelmingly dominating the samples

```{r, visualiation - bar graph, fig.height = 3, fig.width = 4}
# generate a bar graph
ggplot(data = Litter, aes(x = functionalGroup)) + 
geom_bar()
```


15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of `dryMass` by `functionalGroup`.

```{r, visualization - boxplot, violin}
# generate a box-and-whisker plot
ggplot(Litter) + 
geom_boxplot(aes(x = functionalGroup, y = dryMass)) 

# generate a violin plot
ggplot(Litter) + 
geom_violin(aes(x = functionalGroup, y = dryMass),  
draw_quantiles = c(0.25, 0.5, 0.75))

```

► Why is the boxplot a more effective visualization option than the violin plot in this case?
> Answer: The boxplot is more effective because the dryMass data are highly skewed and contain many zero or near-zero values. On the contrary, violin plots rely on density estimation that performs poorly under these conditions.


► What type(s) of litter tend to have the highest biomass at these sites?
> Answer: Needles tend to have the highest biomass at these sites, followed by mixed litter and twigs/branches, which also show relatively high dry mass compared to other litter types.
