---
title: "Assignment 10: Data Scraping"
author: "Erin Lee"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 

library(tidyverse)
library(rvest)
library(here)

here()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 

LWSP_web <- read_html(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
LWSP_web

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 

WaterSystemName <- LWSP_web %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
WaterSystemName

ContactPerson <- LWSP_web %>%
  html_nodes("div+ table tr:nth-child(4) td:nth-child(2)") %>%
  html_text()
ContactPerson

Ownership <- LWSP_web %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
Ownership

MaxDayUse <- LWSP_web %>%
  html_nodes("th~ td+ td") %>%
  html_text()
MaxDayUse

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 

df_LWSP <- data.frame(
  "WaterSystemName" = rep(WaterSystemName, 12), 
  "ContactPerson" = rep(ContactPerson, 12),
  "Ownership" = rep(Ownership, 12),
  "Month" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov",
              "Apr", "Aug", "Dec"),
  "MaxDayUse" = as.numeric(MaxDayUse),
  "Year" = rep(2024, 12)) %>%
  mutate(Date = my(paste(Month,"-",Year)))

df_LWSP  

#5 
ggplot(df_LWSP,aes(x=Date,y=MaxDayUse)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(title = paste("2024 Water usage data for", WaterSystemName),
       y="MaxDayUse_MGD",
       x="Month") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 

scrape.it <- function(year, PWSID){
  
  the_website <- 
    read_html(paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", 
                                  PWSID, "&year=", year))
  
  WaterSystemName_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  ContactPerson_tag <- 'div+ table tr:nth-child(4) td:nth-child(2)'
  Ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  MaxDayUse_tag <- 'th~ td+ td'

  WaterSystemName<- the_website %>% html_nodes(WaterSystemName_tag) %>% html_text()
  ContactPerson <- the_website %>%   html_nodes(ContactPerson_tag) %>%  html_text()
  Ownership <- the_website %>% html_nodes(Ownership_tag) %>% html_text()
  MaxDayUse <- the_website %>% html_nodes(MaxDayUse_tag) %>% html_text()
  
  df_LWSP_2 <- data.frame("Month" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct",
                                      "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"), 
                               "Year" = rep(year,12),
                               "MaxDayUse" = as.numeric(MaxDayUse)) %>% 
    mutate(WaterSystemName = WaterSystemName,
           ContactPerson = ContactPerson,
           Ownership = Ownership,
           PWSID = PWSID,
           Date = my(paste(Month,"-",Year)))

  return(df_LWSP_2)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 

Durham_2020 <- scrape.it(2020,'03-32-010')

ggplot(Durham_2020,aes(x=Date,y=MaxDayUse)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(title = paste("2020 Water usage data for", WaterSystemName),
       y="MaxDayUse_MGD",
       x="Month") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

glimpse(Durham_2020)



```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 

Cary_2020 <- scrape.it(2020, '03-92-020')

combined_2020 <- rbind(Durham_2020, Cary_2020)

ggplot(combined_2020, aes(x = Date, y = MaxDayUse, color = WaterSystemName, 
                          group = WaterSystemName)) +
  geom_line(linewidth = 1) + 
  geom_point() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(
    title = "Comparison of Max Daily Water Use (2020)",
    subtitle = "Durham (03-32-010) vs. Cary (03-92-020)",
    x = "Month",
    y = "MaxDayUse_MGD",
    color = "WaterSystemName"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom" 
  )

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 

the_years = rep(2018:2023)
my_watersystem = '03-92-020'

cary_multi_year <- map2(the_years, my_watersystem, scrape.it) %>% 
  bind_rows()

ggplot(cary_multi_year,aes(x=Date,y=MaxDayUse)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
   title = paste(the_years[1], "-", the_years[length(the_years)], 
                  "Water usage data for Cary"),
    y="MaxDayUse_MGD",
    x="Date"
    ) + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))


```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: Based on the visual inspection of the LOESS smoothed line, Cary shows a slight upward trend in maximum daily water usage from 2018 to 2023. Additionally, the plot exhibits significant seasonal fluctuations, with consistent annual peaks in water demand.
>
